---
title: "Basketball"
author: "Son Luong'code"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(GGally)
library(ggplot2)
library(corrplot)
library(RColorBrewer)
library(stringr)
library(readxl)
library(ggrepel)
```


```{r}
bb_dat <- read.csv('CollegeBasketballPlayers2009-2021.csv')
rank_data <- read_excel("bb_school_ranks.xlsx")
# full_data after cleaning the unusefull variables 
bb_dat <- bb_dat %>% select(-c(27,28,34,35,38,40,41,42,44,45,46,66)) %>% filter(year!=2009) %>% rename (college_year = yr, role = X) 
# %>% rename_with(~str_replace(., '_per',' %'))
top_25s = right_join(bb_dat,rank_data) %>% relocate(college_year , year, role,rank, .before = GP)
```

'porpag' 
'dporpag'
eFG_per:  Effective field goal %: captures scoring efficiency per field goal attempt
TS_per : True shooting % capture the effect of shooting free throws
Box Plus Minus
Minutes played [MP];
Team minutes played [TeMP];
Game played [GP];
Offensive Rebound Percentage [OR%];
Defensive Rebound Percentage [DR%];
Total Rebound Percentage [TR%];
Steal Percentage [ST%];
Block Percentage [BLK%];
Assist Percentage [Ast%];
Usage Percentage [Usg%];
Turnover Percentage/Ratio [TO%];
rimmade: Rim shots made
midmade: Midrange shots made
dunksmade: Dunks made


First , we take a look at how's relationship between all the varibles?
```{r}
numeric_dat <- select_if(bb_dat,is.numeric) %>% select(-contains("per"))%>% na.omit()
corrplot(cor(numeric_dat), type="upper", order="hclust",
         col=brewer.pal(n=8, name="RdYlBu"))

```
So, we see a couples negative correlation dbpm/dgbpm/bpm vs drtg, and pfr vs drtg
postive corralation porpag vs adjoe & obpm, 
pts vs dreb/dporpag/rimmade/midmade/twoPM/FTM/mp/stops
rimmade/dreb/treb/pts
GP vs stops
stl vs mp



```{r}
bb_dat$role[bb_dat$role == ""] <- NA
bb_dat %>% na.omit()%>% filter(rimmade > 100) %>% group_by(role)%>% ggplot( aes(x= rimmade, y = pts, group = role)) + 
  geom_point(aes(col=role))


```


```{r}
bb_dat $pts[bb_dat$pts == ""] <- NA
bb_dat $role[bb_dat$role == ""] <- NA

bb_dat %>% filter(college_year != "Sr") %>% add_count(player_name) %>% filter(n == 10) %>% na.omit() %>% ggplot(aes(x=pts,y=obpm))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + ylab(" Offensive Box Plus/Minus")

#ggplot(aes(x = player,y = TPA, color=player)) + geom_point()

```

How each position attribute into gaining points
```{r}
bb_dat$role[bb_dat$role == ""] <- NA
bb_dat %>% na.omit() %>% ggplot(aes(x=role,y=pts,fill=role)) +
  geom_boxplot(aes(na.rm=T)) + theme_linedraw() + 
  xlab("Role") + ylab("Points")

bb_dat  %>% na.omit() %>% ggplot(aes(x=role,y=pts,fill=role)) + 
  geom_boxplot(aes(x=role,y=pts,fill=role,na.rm=T)) + 
  geom_text_repel(data = subset(bb_dat, pts >= 26), aes(label = player_name))+theme_linedraw() +
  xlab("Role") + 
  ylab("Points")  

```

AST % vs eFG by college year. Should we consider Freshman and Sophomore rather than Junior and Senior fro a dream team? 

> for the AST % vs eFG by college year in top 25 teams, if we look at the upper right quadrant where players with high eFG and AST % distributed, we can see most outlier points in this quadrant are mostly freshman and sophomore.

```{r}
# Top 50 Assistant percentage  vs college year

bb_dat %>% slice_max(AST_per, n=50) %>% count(college_year) %>% filter(college_year == "Fr"|college_year =="So"|college_year =="Jr"|college_year =="Sr") %>% ggplot(aes(x=college_year,y = n,fill = college_year)) + geom_col()+ theme_linedraw() + 
  xlab("College Year") + 
  ylab("Count")

# Top 50 Effective Field Goal vs college year
bb_dat %>% slice_max(eFG, n=50) %>% count(college_year) %>% filter(college_year == "Fr"|college_year =="So"|college_year =="Jr"|college_year =="Sr") %>% ggplot(aes(x=college_year,y = n,fill = college_year)) + geom_col()+ theme_linedraw() + 
  xlab("College Year") + 
  ylab("Count")

# AST % vs eFG by college year

top_25s %>% group_by(player_name) %>% filter(college_year == "Fr"|college_year =="So"|college_year =="Jr"|college_year =="Sr", eFG !=0, AST_per!=0 ) %>% ggplot(aes(x=AST_per,y = eFG)) +
  geom_point(aes(color= college_year))+
  theme_linedraw() +
  xlab("Assistant Percentage") + 
  ylab("Effective Field Goal")
```

Who is the best offensive player?

```{r}
bb_dat $pts[bb_dat$pts == ""] <- NA

#Relationship between Offensive Box Plus/minus vs points
bb_dat $pts[bb_dat$pts == ""] <- NA
bb_dat %>% na.omit() %>% ggplot(aes(x=pts,y=obpm))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + ylab(" Offensive Box Plus/Minus")


# How is the relationship between assistance percentage with Guard Positions
bb_dat %>% ggplot(aes(x=role,y=AST_per))+ 
  geom_col() +  
  theme_minimal()+
  xlab("Player Role") + 
  ylab(" Accumulated Assistent Percentage")


# Offensive Rating vs Points among players have 100 % assistance percentage
bb_dat %>% filter(AST_per == 100, Ortg!= 0, pts!=0) %>% 
  ggplot(aes(x= Ortg, y= pts))+ 
  geom_point() + 
  geom_text (hjust = 0, nudge_x = 0.05,aes(label= player_name, colour = role))+
  theme_light()+
  xlab("Offensive Rate") + 
  ylab("Points")
```



```{r}


```


The more they contribute on the court, the more points they get ?
```{r}
bb_dat $pts[bb_dat$pts == ""] <- NA
bb_dat %>% na.omit() %>% ggplot(aes(x=pts,y=bpm))+
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + 
  ylab("Box Plus/Minus")
```
The relationship between Effective Field Goal vs Box Plus/Minus

```{r}
bb_dat$eFG[bb_dat$eFG == ""] <- NA
bb_dat %>% na.omit() %>% ggplot(aes(x=eFG,y=bpm))+
  geom_smooth(se = F) + 
  theme_light() + 
  xlab("Effective Field Goal") + 
  ylab("Box Plus/Minus")
```

The relationship Game played and Effective Field Goal?
Especially, Power Forward Position seem like  experience is very important to consider selecting players for this position in the dream team for a higher effective field goal
```{r}
bb_dat$eFG[bb_dat$eFG == ""] <- NA

bb_dat %>%  na.omit() %>% ggplot(aes(y=eFG,x=GP))+
  geom_smooth(aes(colour= role), se=F) +
  theme_minimal() +
  xlab("Game Glayed") +
  ylab("Effective Field Goal")
```

Point vs Assistance by Team in top    10
```{r}
top_25s %>%na.omit() %>% group_by(team) %>%
summarize(n = sum(ast), m = sum(pts)) %>%
head(10) %>%  ggplot(aes(x=n,y=m))+ 
  geom_point(aes(col = team)) +  
  geom_text(aes(label=team,col=team))+
  xlab("Assistance") +
  ylab("Points")

bb_dat %>%na.omit() %>% group_by(team) %>%
summarize(n = sum(ast), m = sum(pts)) %>%
head(10) %>%  ggplot(aes(x=n,y=m))+ 
  geom_point(aes(col = team)) +  
  geom_text(aes(label=team,col=team))+
  xlab("Assistance") +
  ylab("Points")
```

