---
title: "Basketball"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(GGally)
library(ggplot2)
library(corrplot)
library(RColorBrewer)
library(stringr)

```


```{r}
data <- read.csv('Basketball.csv')
str(data)
```



Split the data into several data frame based on variable's types for different use for later analysis and visualization: continuous data and categorical data, full_data after cleaning the unusefull variables and missing values


```{r}

# full_data after cleaning the unusefull variables and missing values
data <- data %>% relocate(college_year = yr, year, role= X, .before = GP) %>% filter(year!=2009) %>%  na.omit()

```


```{r}
# numeric data
numeric_data <- data %>% select(-c(1:6,29,30,35,39,41:43,45,46))
#percentage data
per_data <- numeric_data %>% select(contains("per")) %>% rename_with(~str_replace(., '_per',' %'))
#continuous data
continuous_data <- numeric_data%>% select(-contains("per"))
#categorical data
categorical_data <- data[,1:6]
```

Use correlation plot to see how all continuous variables correlated to each other

```{r}
corrplot(cor(continuous_data), type="upper", order="hclust",
         col=brewer.pal(n=8, name="RdYlBu"))

```

#visualize year of college ~ point
#group role position & year of college
```{r}

```
