---
title: "Basketball"
author: "Son Luong and Will Curkan"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(GGally)
library(ggplot2)
library(corrplot)
library(RColorBrewer)
library(stringr)
library(readxl)
library(ggrepel)
library(ggpubr)
```
## Introduction:

Statisticss are a good way to tell how you are doing in a sport and what you are best at. They can also help to determine where you and your team need to improve. We can use statistic in sport to create a dream team. We work on the College Basketball created by Aditya Kumar from Kaggle, which contains past decade's NCAA college basketball data.

 
```{r}
bb_dat <- read.csv('CollegeBasketballPlayers2009-2021.csv')
rank_data <- read_excel("bb_school_ranks.xlsx")
# full_data after cleaning the unusefull variables 
bb_dat <- bb_dat %>% select(-c(27,28,34,35,38,40,41,42,44,45,46,66)) %>% filter(year!=2009) %>% rename (college_year = yr, role = X) 
# %>% rename_with(~str_replace(., '_per',' %'))
top_25s = right_join(bb_dat,rank_data) %>% relocate(college_year , year, role,rank, .before = GP)
```


## How's relationship between all the varibles?

First , we take a look at how's relationship between all the varibles? 

```{r}
numeric_dat <- select_if(bb_dat,is.numeric) %>% select(-contains("per"))%>% na.omit()
corrplot(cor(numeric_dat), type="upper", order="hclust",
         col=brewer.pal(n=8, name="RdYlBu"))

```
So, we notice a couples negative correlation dbpm/dgbpm/bpm vs drtg, and pfr vs drtg and some postive correlation porpag vs adjoe & obpm, pts vs dreb/dporpag/rimmade/midmade/twoPM/FTM/mp/stops, rimmade/dreb/treb/pts, GP vs stops, stl vs mp. 





```{r}
bb_dat$role[bb_dat$role == ""] <- NA
bb_dat %>% na.omit()%>% filter(rimmade > 100) %>% group_by(role)%>% ggplot( aes(x= rimmade, y = pts, group = role)) +
  geom_point(aes(col=role, alpha=0.01))+
  facet_wrap(~role)


```

Sports is a teamwork!Then how to evaluate a good team player, we measure player’s contributions and their impact on the court by BPM (Box Plus/Minus). The more time they contribute on the court, the more points they get. 

```{r}
bb_dat $pts[bb_dat$pts == ""] <- NA
bb_dat $role[bb_dat$role == ""] <- NA
bb_dat $mp[bb_dat$mp == ""] <- NA

bb_dat %>% filter(college_year != "Sr") %>% add_count(player_name) %>% filter(n == 10) %>% na.omit() %>% ggplot(aes(x=pts,y=bpm))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + ylab(" Box Plus/Minus")+ ggtitle("Box Plus/Minus vs. Points")

bb_dat %>% filter(college_year != "Sr") %>% add_count(player_name) %>% filter(n == 10) %>% na.omit() %>% ggplot(aes(x=pts,y=mp))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + ylab(" Minutes Played")+ ggtitle("Minutes Played vs. Points")

bb_dat %>% filter(college_year != "Sr") %>% add_count(player_name) %>% filter(n == 10) %>% na.omit() %>% ggplot(aes(x=bpm,y=mp))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Box Plus Minus") + ylab(" Minutes Played") + ggtitle("Minutes Played vs. Box Plus/Minus")


#ggplot(aes(x = player,y = TPA, color=player)) + geom_point()

```

How each position attribute into gaining points
```{r}
bb_dat$role[bb_dat$role == ""] <- NA
bb_dat %>% na.omit() %>% ggplot(aes(x=role,y=pts,fill=role)) +
  geom_boxplot(aes(na.rm=T)) + theme_linedraw() + 
  xlab("Role") + ylab("Points")

bb_dat%>% na.omit() %>% ggplot(aes(x=role,y=pts,fill=role)) + 
  geom_boxplot(aes(x=role,y=pts,fill=role,na.rm=T)) + 
  geom_text_repel(data = filter(bb_dat, pts >= 23), aes(label = player_name,col=role)) + 
  theme_linedraw() +
  xlab("Role") + 
  ylab("Points")  

```
## How  academic status (freshman, sophmores, juniors, and seniors) affect their performance?

AST % vs eFG by college year. Should we consider Freshman and Sophomore rather than Junior and Senior? 

> for the AST % vs eFG by college year in top 25 teams, if we look at the upper right quadrant where players with high eFG and AST % distributed, we can see most outlier points in this quadrant are mostly freshman and sophomore.

```{r}
# Top 50 Assistant percentage  vs college year

a<- bb_dat %>% slice_max(AST_per, n=100) %>% count(college_year) %>% filter(college_year == "Fr"|college_year =="So"|college_year =="Jr"|college_year =="Sr") %>% mutate(college_year = recode(college_year, "Fr"= "Freshman","So" = "Sorphomore", "Jr" = "Junior", "Sr" = "Senior")) %>% ggplot(aes(x=college_year,y = n,fill = college_year)) + geom_col()+ 
  labs(fill = "College Year\n")+
  xlab("College Year") + 
  ylab("Count")+ ggtitle("Top 50 Assistant Percentage by College Year")+theme_linedraw() 

# Top 50 Effective Field Goal vs college year
b<-bb_dat %>% slice_max(eFG, n=100) %>% count(college_year) %>% filter(college_year == "Fr"|college_year =="So"|college_year =="Jr"|college_year =="Sr")  %>% mutate(college_year = recode(college_year, "Fr"= "Freshman","So" = "Sorphomore", "Jr" = "Junior", "Sr" = "Senior")) %>% ggplot(aes(x=college_year,y = n,fill = college_year)) + geom_col()+ theme_linedraw() + 
  labs(fill = "College Year\n")+
  xlab("College Year") + 
  ylab("Count") + ggtitle("Top 50 Effective Field Goal by College Year")

# AST % vs eFG by college year

c <- top_25s %>% group_by(player_name) %>% filter(college_year == "Fr"|college_year =="So"|college_year =="Jr"|college_year =="Sr", eFG !=0, AST_per!=0 ) %>% ggplot(aes(x=AST_per,y = eFG)) +
  geom_point(aes(color = college_year))+
  facet_wrap(~college_year)+
  labs(color = "College Year\n") +
  xlab("Assistant Percentage")+
  ylab("Effective Field Goal %") + 
  ggtitle("Assistant Percentage vs. Effective Field Goal % \n by college year") + 
  theme_minimal() 
ggarrange(a,b,c, ncol = 2, nrow = 2)

```

## The more they contribute on the court, the more points they get ?
```{r}
bb_dat $pts[bb_dat$pts == ""] <- NA
bb_dat %>% na.omit() %>% ggplot(aes(x=pts,y=bpm))+
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + 
  ylab("Box Plus/Minus")+ ggtitle("Relationship between  Box Plus Minus and Points")
```
## Who is the best offensive players?

```{r}
bb_dat $pts[bb_dat$pts == ""] <- NA

#Relationship between Offensive Box Plus/minus vs points
bb_dat $pts[bb_dat$pts == ""] <- NA
bb_dat %>% na.omit() %>% ggplot(aes(x=pts,y=obpm))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + ylab(" Offensive Box Plus/Minus")+ ggtitle("Relationship between Offensive Box Plus Minus and Points")


# Offensive Rating vs Points among players have 100 % assistance percentage
bb_dat %>% filter(AST_per == 100, Ortg!= 0, pts!=0) %>% 
  ggplot(aes(x= Ortg, y= pts))+ 
  geom_point(aes(color=role)) + 
  geom_text_repel(hjust = 0, nudge_x = 0, 
                  aes(label= player_name, colour = role),size=2)+  
  theme_minimal()+
  xlab("Offensive Rating") + ylab("Points") + ggtitle("Offensive Rating vs. Points \n Among Players with 100 % assistance percentage")
```
## Who is best defensive player who are 
Playing great defense will lead to more offense, minimize opponents possession time and shooting percentage per possession.
Defensive Rating (DRtg)
Defensive Rebounding Percentage (DRB%)
Block Percentage (BLK%)
Steal Percentage (STL%)

```{r}
bb_dat %>% select(1:2,role, drtg, blk_per,stl_per,DRB_per) %>% na.omit() %>% group_by(role)%>% ggplot( aes(x= drtg, y = blk_per, group = role)) + geom_point(aes(col=role, alpha=0.01))+ facet_wrap(~role)
```



## Who is the best Shooters? 

```{r}

```


The relationship between Effective Field Goal vs Box Plus/Minus

```{r}
bb_dat$eFG[bb_dat$eFG == ""] <- NA
bb_dat %>% na.omit() %>% ggplot(aes(x=eFG,y=bpm))+
  geom_smooth(se = F) + 
  theme_light() + 
  xlab("Effective Field Goal") + 
  ylab("Box Plus/Minus")

```

The relationship Game played and Effective Field Goal?
Especially, Power Forward Position seem like  experience is very important to consider selecting players for this position in the dream team for a higher effective field goal
```{r}
bb_dat$eFG[bb_dat$eFG == ""] <- NA
quantile(bb_dat$eFG,c(.25,.5,.75))

bb_dat %>%  na.omit() %>% ggplot(aes(y=eFG,x=GP))+
  geom_smooth(aes(colour = role), se=F) +
  theme_classic() +
  facet_wrap(~role)+
  labs(x="Game Glayed", y="Effective Field Goal %", title="Effective Field Goal % vs. Game played \nby Role")
```
  
Point vs Assists by Team in top 10 team have highest accummulated 
```{r}
top_25s %>%na.omit() %>% group_by(team) %>%
summarize(n = sum(ast), m = sum(pts)) %>%
head(10) %>%  ggplot(aes(x=n,y=m))+ 
  geom_point(aes(col = team)) +  
  geom_text(aes(label=team,col=team))+
  xlab("Assists") +
  ylab("Points")

bb_dat %>%na.omit() %>% group_by(team) %>%
summarize(ast_sum = sum(ast), m = sum(pts)) %>%
slice_max(ast_sum, n = 10) %>%  ggplot(aes(x=ast_sum,y=m))+ 
  geom_point(aes(col = team)) +  
  geom_text_repel(aes(label=team,col=team))+
  labs(x="Assistance", y="Points", title="Assistance vs Points \nTop 10 Team with Highest Assistance")
  
```


```{r free throws}
f1<- bb_dat %>%
  filter(FTA > 58 & FT_per > .769) %>%
  ggplot(aes(x = FTA,y = FT_per)) +
  geom_point() + 
  geom_smooth(se=F)+
  xlab("Free Throw Attempts") +
  ylab("Free Throw Percentage")+
  ggtitle("General Free Throw Comparison")
#guards
f2 <- bb_dat %>%
  filter(FTA > 58 & FT_per > .769) %>%
      filter(role == "Combo G" | role == "Scoring PG"
        | role == "Wing G" | role == "Pure PG") %>%
  ggplot(aes(x = FTA,y = FT_per)) +
  geom_point() + 
  geom_smooth(se=F)+
  xlab("Free Throw Attempts") +
  ylab("Free Throw Percentage")+
  ggtitle("Guard Free Throws")

#Centers/Powerforwards
f3 <- bb_dat %>%
  filter(FTA > 58 & FT_per > .769) %>%
  filter(role == "PF/C" | role == "C"
        | role == "Wing F" | role == "Stretch 4") %>%
  ggplot(aes(x = FTA,y = FT_per)) +
  geom_point() + 
  geom_smooth(se=F)+
  xlab("Free Throw Attempts") +
  ylab("Free Throw Percentage")+
  ggtitle("Post Player Free Throws")
#quantile(bb_dat$FT_per,c(.25,.5,.75))+
#quantile(bb_dat$FTA,c(.25,.5,.75))
ggarrange(f1,f2,f3, ncol = 3, nrow = 1)
```
```{r rebound comparison}
r1<-bb_dat %>%
  filter(GP > 29) %>%
  ggplot(aes(x = ORB_per,y = DRB_per)) +
  geom_point() + 
  geom_smooth()+ 
  xlab("Offensive Rebound Percentage")+
  ylab("Defensive Rebound Percentage")+
  ggtitle("Rebound Comparison")
#Guards
r2<-bb_dat %>%
  filter(GP > 29) %>%
    filter(role == "Combo G" | role == "Scoring PG"
        | role == "Wing G" | role == "Pure PG") %>%
  ggplot(aes(x = ORB_per,y = DRB_per)) +
  geom_point() + 
  geom_smooth()+ 
  xlab("Offensive Rebound Percentage")+
  ylab("Defensive Rebound Percentage")+
  ggtitle("Guard Rebound Comparison")
#Center/Powerforwards
r3<-bb_dat %>%
  filter(GP > 29) %>%
  filter(role == "PF/ C" | role == "C"| role == "Wing F" | role == "Stretch 4") %>%
  ggplot(aes(x = ORB_per,y = DRB_per)) +
  geom_point() + 
  geom_smooth()+ 
  xlab("Offensive Rebound Percentage")+
  ylab("Defensive Rebound Percentage")+
  ggtitle("Post Player Rebound Comparison")

  #quantile(bb_dat$GP,c(.25,.5,.75))


#r4<- bb_dat %>% filter(GP > 29)  %>% ggplot(aes(y=ORB_per,x=DRB_per))+ geom_smooth(aes(colour = role), se=F) + theme_classic() + facet_wrap(~role)+ labs(x="DRB", y="ORB", title="Defensive Rebound  vs. Offensive Rebound \nby Role")

ggarrange(r1,r2,r3, ncol = 3, nrow = 1)

```
##The assist to turnover ratio 
The assist to turnover ratio is one of many basketball statistics used to evaluate the ball control and ball handling skills of a player
```{r}
bb_dat %>% ggplot(aes(x=ast.tov,y=Ortg))+
  geom_smooth(aes(colour = role), se=F) +
  theme_classic() +
  facet_wrap(~role)+
  labs(x="The assist to turnover ratio", y="Offensive Rating", title="The assist to turnover ratio vs.Offensive Rating \nby Role")

bb_dat %>% ggplot(aes(x=ast.tov,y=drtg))+
  geom_smooth(aes(colour = role), se=F) +
  theme_classic() +
  facet_wrap(~role)+
  labs(x="The assist to turnover ratio", y="Defensive Rating", title="The assist to turnover ratio vs. Defensive Rating \nby Role")
bb_dat %>% ggplot(aes(x=ast.tov,y=AST_per))+
  geom_smooth(aes(colour = role), se=F) +
  theme_classic() +
  facet_wrap(~role)+
  labs(x="The assist to turnover ratio", y="Assist Percentage", title="The assist to turnover ratio vs. Assist Percentage \nby Role")
```




The data is the  filtered down to the last 3 seasons 18-19,19-20,20-21 for simplicity. 
```{r top 10 asts}
options(scipen = 999)
top_25s %>%
  na.omit() %>%
  filter(appearances == 3) %>%
  group_by(team) %>%
  summarize(n = sum(ast))%>%
  arrange(desc(n)) %>%
  ggplot()+
  geom_col(aes(x = team,y = n,fill = team))+
  ggtitle("Total Assist for Team with 3 appearances")+
  xlab("Team")+
  ylab("Assists")+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())

top_25s %>%
  na.omit() %>%
  filter(appearances == 2)%>%
  group_by(team) %>%
  summarize(n = sum(ast)) %>%
  arrange(desc(n)) %>%
  ggplot()+
  geom_col(aes(x = team,y = n,fill = team))+
  ggtitle("Total Assist for Team with 2 appearances")+
  xlab("Team")+
  ylab("Assists")+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())

top_25s %>%
  na.omit() %>%
  filter(appearances == 1) %>%
  group_by(team) %>%
  summarize(n = sum(ast))%>%
  arrange(desc(n)) %>%
  ggplot()+
  geom_density(aes(x = n,fill = team))+
  ggtitle("Total Assist for Team With 1 Appearance")+
  xlab("Team")+
  ylab("Assists")+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
#full graph with all 25 teams
ast_and_team = top_25s %>%
  na.omit() %>%
  group_by(team) %>%
  summarize(n = mean(ast))%>%
  arrange(desc(n)) 

ast_Nteam_rank = full_join(ast_and_team,new_ranks) %>%
  ggplot() +
  geom_density(aes(x = n, color = as.factor(appearances)))
ast_Nteam_rank
  
```

```{r}
ast_and_team = top_25s %>%
  na.omit() %>%
  group_by(school) %>%
  summarize(mean_ast = mean(ast))
rank_data %>% add_count(school)%>% rename(appearances = n) %>% full_join(ast_and_team) %>%
  ggplot() +
  geom_density(aes(x = mean_ast, color = as.factor(appearances)))

```

```{r ast/tov}
top_25s %>%
  na.omit() %>%
  filter(ast.tov > 1 & GP > 30) %>%
  ggplot(aes(x = GP,y = ast.tov))+
  geom_point()+
  geom_smooth()
  
```

In order to create a dream team for the 2022 season we must look at freshman in 2018-2019, freshman and sophmores in 2019-2020, and freshman, sophmore, and juniors in 2020-2021
```{r freshman 2019 filters}
#top freshman guards
top_fr_guards = top_25s %>%
  filter(college_year == 'Fr' & year == 2019 & GP > 25) %>%
  filter(role == "Combo G" | role == "Scoring PG"
        | role == "Wing G" | role == "Pure PG")

#top powerforwards and centers
top_fr_PfC = top_25s %>%
  filter(college_year == 'Fr' & year == 2019) %>%
  filter(role == "PF/C" | role == "C"
        | role == "Wing F" | role == "Stretch 4")
# corrplot(cor(cont_dat),type="upper", order="hclust",col=brewer.pal(n=8, name="RdYlBu"))
``` 



```{r freshman plots}
top_fr_guards %>% 
  select(where(is.double))
top_fr_guards %>%
  ggplot(aes(x = twoPA)) +
  geom_histogram()
top_fr_PfC %>%
  ggplot(aes(x = twoPA,y = twoPM)) +
  geom_violin()
top_fr_PfC %>%
  arrange(desc(DRB_per,twoP_per)) %>%
  slice(1:5) %>%
  ggplot(aes(x = DRB_per,y = twoP_per)) +
  geom_point() +
  geom_smooth() + 
  geom_text(aes(label=player_name))
top_fr_PfC %>%
  arrange(desc(DRB_per,twoP_per)) %>%
  slice(1:5) %>%
  ggplot(aes(x = DRB_per,y = twoP_per)) +
  geom_point() +
  geom_smooth() + 
  geom_text(aes(label=player_name))
#filter for top DRB percentage and shooting
top_fr_PfC %>%
  arrange(desc(DRB_per,twoP_per)) %>%
  filter(Min_per > 50)
  
```



```{r fresh_soph_2020 filters}
#filters for freshman and sophmores in 2018-2019 season
top_25_2soph_2020 = top_25s %>%
  filter(college_year == "Fr" & year == 2020
         |college_year == "So" & year == 2020)
#top guards
top_2soph_role_g_2020 = top_25_2soph_2020 %>%
    filter(role == "Combo G" | role == "Scoring PG"
         | role == "Wing G" | role == "Pure PG")
#top powerforwards/center
top_2soph_role_pfC_2020 = top_25_2soph_2020 %>%
      filter(role == "PF/C" | role == "C"
         | role == "Wing F" | role == "Stretch 4")
```



```{r to juniors 2021}
#freshman, sophmore, juniors in 2020-2021 season
#top guards
top_2jnr_guards = top_25s %>%
  filter(!yr == "Sr") %>%
  filter(role == "Combo G" | role == "Scoring PG"
         | role == "Wing G" | role == "Pure PG")
#top powerforwards/center
top_2jnr_PfC = top_25s %>%
  filter(!yr == "Sr") %>%
  filter(role == "PF/C" | role == "C"
         | role == "Wing F" | role == "Stretch 4")
```




1. Correlation between year (freshman, sophmore, etc) and stats
or recurring players and juniors


```{r}
#looking at individual recurring players stats
recurring_andJuniors = top_25s %>%
  add_count(player_name) %>%
  filter(college_year == "Jr" | n == 3 & GP > 20)
#use scatterplot to compare individual recurring stats maybe
recurring_players %>%
  filter(GP > 20) %>%
  ggplot(aes(x=player_name,y=pts)) + 
  geom_point(aes(x=player_name,y=pts,color=player_name)) + 
  theme_linedraw() +
  xlab("Role") + 
  ylab("Points") +
  scale_x_discrete(labels = paste0('p',c(1:9)))
```






```{r}
new_bb$role[new_bb$role == ""] <- NA
new_bb %>% na.omit() %>% ggplot(aes(x=role,y=pts,fill=role)) +
  geom_boxplot(aes(na.rm=T)) + theme_linedraw() + 
  xlab("Role") + ylab("Points")
new_bb  %>% na.omit() %>% ggplot(aes(x=role,y=pts,fill=role)) + 
  geom_boxplot(aes(x=role,y=pts,fill=role,na.rm=T)) + 
  geom_text_repel(data = subset(bb_dat, pts >= 26), aes(label = player_name))+theme_linedraw() +
  xlab("Role") + 
  ylab("Points") 
new_bb %>% slice_max(AST_per, n=50) %>% count(yr) %>% filter(yr == "Fr"|yr =="So"|yr =="Jr"|yr =="Sr") %>% ggplot(aes(x=yr,y = n,fill = yr)) + geom_col()+ theme_linedraw() + 
  xlab("College Year") + 
  ylab("Count")
```


```{r freshman plots}
#looking at freshman playing most games and scoring highest
top_25_fr_2019 %>%
  filter(GP > 20 & Min_per > 50) %>%
  ggplot(aes(x = twoP_per,y = TP_per)) + 
  geom_point() + 
  geom_text(aes(label = player_name),size = 4,ljust = .25)

top_fr_role_g %>%
  filter(Min_per > 59.9 & AST_per > 20 ) %>%
  ggplot(aes(x = twoP_per,y = TP_per)) + 
  geom_point() + 
  geom_text(aes(label = player),size = 4,ljust = .25)
```

3. Correlation between position and points scored(barplot) group-by(team), facet year
4. Correlation between rebounds(offense and def) position, and points scored (continuous)
```{r}
top_25s %>%
filter(year == 2020) %>%
  ggplot(aes(x = team,y = FTA)) +
  geom_bar()
  
```

5. Look at averages for all players and see which freshman stick out
6. filter players not appearing in latter years
7. Look at who improves throughout the years
filter for players who only play to junior year