---
title: "What It Takes"
subtitle: "to Play NCAA College Basketball"
author: "Will Curkan and Son Luong"
date: "12/6/2021"
output:
  slidy_presentation: 
    fig_height: 3
    fig_width: 3
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = F, warning = F, echo = F}
library(tidyverse)
library(dplyr)
library(GGally)
library(ggplot2)
library(corrplot)
library(RColorBrewer)
library(stringr)
library(readxl)
library(ggrepel)
library(ggpubr)
```

## Introduction and Exploring Data

Statistics are a good way to tell how you are doing in a sport and what you are best at. They can also help to determine where you and your team need to improve. We can use statistic in sport to create a dream team. We work on the College Basketball created by Aditya Kumar from Kaggle, which contains past decade's NCAA college basketball data.

Our goal was to look at data trends based on role and academic status: freshman, sophmores, juniors, and seniors. The data is the past decade's NCAA college basketball.


---

```{r import data, echo = F}
bb_dat = read.csv("CollegeBasketballPlayers2009-2021.csv")
school_dat = read_excel("bb_school_ranks.xlsx")

```


Cleaned data to make easier to work with:

- rename variables to readable, workable names
- filter to recent 3 years
- remove less-meaningful variables

```{r data cleanup, echo = F, message = F}
new_bb = bb_dat %>%
  rename(role = X, player = player_name) %>%
  filter(year > 2018) %>%
  select(!ht & !dunksmade & !dunksmiss.dunksmade
         & !dunksmade..dunksmade.dunksmiss.
         & !type & !num & !pid & !pick & !pfr
         & !Rec.Rank)

son_bb_dat <- bb_dat %>% select(-c(27,28,34,35,38,40,41,42,44,45,46,66)) %>% filter(year!=2009) %>% rename (college_year = yr, role = X)

#rename school to team to match new_bb
new_ranks = school_dat %>%
  rename(team = school) %>%
  add_count(team)%>%
  rename(appearances = n)

#creating vector for top team in whole dataset
top_teams_vector = new_ranks$team

#recreating new_bb filter to show if player in top team
new_bb = new_bb %>%
  mutate(top_team = ifelse(team %in% top_teams_vector,T,F))

#46 teams have been in top 25 from (2018-)2019-2021
top_25s = inner_join(new_bb,new_ranks) %>%
  relocate(college_year = yr, year, role, .before = GP)
```


## Fun Fax

- How many teams are represented in this dataset

```{r echo = F}
#unique teams in NCAA n = 358, all divisions
new_bb %>%
  distinct(team) %>%
  count() %>%
  rename(unique_teams = n)
```


- How many unique players in NCAA 2018+

```{r echo = F}
new_bb %>%
  distinct(player) %>%
  count() %>%
  rename(unique_players = n)
```

- How many recurring players from seasons 2018-19+
```{r echo = F}
top_25s %>%
  filter(!college_year == "Sr") %>%
  add_count(player) %>%
  filter(n == 3) %>%
  group_by(player) %>%
  summarize(n = n()) %>%
  count() %>%
  rename(recurring_players = n)
```


---

How's relationship between all the varibles?

First, we take a look at how's relationship between all the varibles? 


```{r fig.height = 5, fig.width = 5, echo = F}
numeric_dat <- select_if(son_bb_dat,is.numeric) %>% select(-contains("per"))%>% na.omit()
corrplot(cor(numeric_dat), type="upper", order="hclust",
         col=brewer.pal(n=8, name="RdYlBu"))
```


So, we notice a couples negative correlation dbpm/dgbpm/bpm vs drtg, and pfr vs drtg and some postive correlation porpag vs adjoe & obpm, pts vs dreb/dporpag/rimmade/midmade/twoPM/FTM/mp/stops, rimmade/dreb/treb/pts, GP vs stops, stl vs mp. 


## Visualization

Sports require teamwork! For how to evaluate a good team player, we measure player’s contributions and their impact on the court by BPM (Box Plus/Minus). The more time they contribute on the court, the more points they get. 


```{r echo = F, message = F, warning = F, fig.height = 3, fig.width = 8}
son_bb_dat $pts[son_bb_dat$pts == ""] <- NA
son_bb_dat $role[son_bb_dat$role == ""] <- NA
son_bb_dat $mp[son_bb_dat$mp == ""] <- NA
a = son_bb_dat %>% filter(college_year != "Sr") %>% add_count(player_name) %>% filter(n == 10) %>% na.omit() %>% ggplot(aes(x=pts,y=bpm))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + ylab(" Box Plus/Minus")+ ggtitle("Box Plus/Minus vs. Points")
b = son_bb_dat %>% filter(college_year != "Sr") %>% add_count(player_name) %>% filter(n == 10) %>% na.omit() %>% ggplot(aes(x=pts,y=mp))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Points") + ylab(" Minutes Played")+ ggtitle("Minutes Played vs. Points")
c = son_bb_dat %>% filter(college_year != "Sr") %>% add_count(player_name) %>% filter(n == 10) %>% na.omit() %>% ggplot(aes(x=bpm,y=mp))+ 
  geom_smooth(se = F) +  
  theme_light() + 
  xlab("Box Plus Minus") + ylab(" Minutes Played") + ggtitle("Minutes Played vs. Box Plus/Minus")
ggarrange(a,b,c, ncol = 3, nrow = 1)
```

---

How Each Position Attributes into Gaining Points


```{r message = F, warning = F, echo = F, fig.height= 4, fig.width = 7}
son_bb_dat%>% na.omit() %>% ggplot(aes(x=role,y=pts,fill=role)) + 
  geom_boxplot(aes(x=role,
                   y=pts,
                   fill=role,
                   na.rm=T)) + 
  geom_text_repel(data = filter(son_bb_dat, pts >= 23), 
                  aes(label = player_name,
                      col=role)) + 
  theme_linedraw() +
  xlab("Role") + 
  ylab("Points")  
```


---

How  academic status (freshman, sophmores, juniors, and seniors) affect their performance?

> for the AST % vs eFG by college year in top 25 teams, if we look at the upper right quadrant where players with high eFG and AST % distributed, we can see most outlier points in this quadrant are mostly freshman and sophomore.

```{r echo = F, fig.height = 3, fig.width = 8}
# Top 50 Assistant percentage  vs college year

d<-son_bb_dat %>% slice_max(AST_per, n=100) %>% count(college_year) %>% filter(college_year == "Fr"|college_year =="So"|college_year =="Jr"|college_year =="Sr") %>% mutate(college_year = recode(college_year, "Fr"= "Freshman","So" = "Sorphomore", "Jr" = "Junior", "Sr" = "Senior")) %>% ggplot(aes(x=college_year,y = n,fill = college_year)) + geom_col()+ 
  labs(fill = "College Year\n")+
  xlab("College Year") + 
  ylab("Count")+ ggtitle("Top 50 Assistant Percentage by College Year")+theme_linedraw() 

# Top 50 Effective Field Goal vs college year
e<-son_bb_dat %>% slice_max(eFG, n=100) %>% count(college_year) %>% filter(college_year == "Fr"|college_year =="So"|college_year =="Jr"|college_year =="Sr")  %>% mutate(college_year = recode(college_year, "Fr"= "Freshman","So" = "Sorphomore", "Jr" = "Junior", "Sr" = "Senior")) %>% ggplot(aes(x=college_year,y = n,fill = college_year)) + geom_col()+ theme_linedraw() + 
  labs(fill = "College Year\n")+
  xlab("College Year") + 
  ylab("Count") + ggtitle("Top 50 Effective Field Goal by College Year")

ggarrange(d,e, ncol = 2, nrow = 1)
```


---

The relationship between Game played and Effective Field Goal?
The Powerforward Position seems like  experience is very important to consider.


```{r message = F, warning=F, echo = F, fig.width = 7, fig.height = 3}
son_bb_dat$eFG[son_bb_dat$eFG == ""] <- NA
son_bb_dat %>%  na.omit() %>% ggplot(aes(y=eFG,x=GP))+
  geom_smooth(aes(colour = role), se=F) +
  theme_classic() +
  facet_wrap(~role)+
  labs(x="Game Glayed", y="Effective Field Goal %", title="Effective Field Goal % vs. Game played \nby Role")
```

---

Point vs Assistance by Team in top 10 team have highest accummulated 


```{r echo = F, fig.width = 7, fig.height = 4}

son_bb_dat %>%na.omit() %>%
  group_by(team) %>%
summarize(ast_sum = sum(ast)
          , m = sum(pts)) %>%
slice_max(ast_sum, n = 10) %>%
  ggplot(aes(x=ast_sum,y=m))+ 
  geom_point(aes(col = team)) +  
  geom_text_repel(aes(label=team,col=team))+
  labs(x="Assists", y="Points", title="Assists vs Points \nTop 10 Team with Highest Assistance")
  
son_bb_dat %>%
  group_by(team) %>%
summarize(ast_sum = sum(ast)
          , m = sum(pts)) %>%
  ggplot(aes(x=ast_sum,y=m))+ 
  geom_jitter()+
  labs(x="Assists", y="Points", title="Assists vs Points")
```


## Free Throws

Now we categorize all guard variations as guards, and all the Powerforwards, centers and stretch as post players.

Free Throws are incredibly important because they are uncontested shots 


```{r free throws, echo = F, message = F}
# quantile(new_bb$FT_per,c(.25,.5,.75))
# quantile(new_bb$FTA,c(.25,.5,.75))

new_bb %>%
  # filter(FTA > 58 & FT_per > .769) %>%
  ggplot(aes(x = FTA,y = FT_per)) +
  geom_point() + 
  geom_smooth()+
  xlab("Free Throw Attempts") +
  ylab("Free Throw Percentage")+
  ggtitle(label = "General Free Throw",
          subtitle = "all players")

#guards
new_bb %>%
  filter(FTA > 58 & FT_per > .769) %>%
      filter(role == "Combo G" | role == "Scoring PG"
        | role == "Wing G" | role == "Pure PG") %>%
  ggplot(aes(x = FTA,y = FT_per)) +
  geom_point() + 
  geom_smooth()+
  xlab("Free Throw Attempts") +
  ylab("Free Throw Percentage")+
  ggtitle("Guard Free Throws")

#Centers/Powerforwards
new_bb %>%
  filter(FTA > 58 & FT_per > .769) %>%
  filter(role == "PF/C" | role == "C"
        | role == "Wing F" | role == "Stretch 4") %>%
  ggplot(aes(x = FTA,y = FT_per)) +
  geom_point() + 
  geom_smooth()+
  xlab("Free Throw Attempts") +
  ylab("Free Throw Percentage")+
  ggtitle("Post Player Free Throws")

#guard with most Free throws and highest percentage
#Deishuan Booker
new_bb %>%
  filter(FTA > 250 & FT_per > .9) %>%
  select(player, team, role, TP_per)
```

## Rebounds

Rebounds are important because this is where the teams show who wnats it more, this is where the game get's physical, and personal.

```{r rebound comparison, echo = F, message = F, fig.height = 3, fig.width = 8}
# quantile(new_bb$GP,c(.25,.5,.75))
# quantile(new_bb$ORB_per,c(.25,.5,.75))
# quantile(new_bb$DRB_per,c(.25,.5,.75))

r1 = new_bb %>%
  filter(GP > 29) %>%
  ggplot(aes(x = ORB_per,y = DRB_per)) +
  geom_point() + 
  geom_smooth()+ 
  xlab("Offensive Rebound Percentage")+
  ylab("Defensive Rebound Percentage")+
  ggtitle(label = "Rebound Comparison",
          subtitle = "all players")


#Center/Powerforwards
r2 = new_bb %>%
  filter(GP > 29 & ORB_per >= 7.6 & DRB_per >= 16.7) %>%
  filter(role == "PF/C" | role == "C"
        | role == "Wing F" | role == "Stretch 4") %>%
  filter(!yr == "Sr") %>%
  ggplot(aes(x = ORB_per,y = DRB_per)) +
  geom_point() + 
  geom_smooth()+
  facet_wrap(~yr)+
  xlab("Offensive Rebound Percentage")+
  ylab("Defensive Rebound Percentage")+
  labs(title = "Post Player Rebounds",
       subtitle = "Q3+")


#Guards
new_bb %>%
  filter(GP > 29 & ORB_per >= 7.6 & DRB_per >= 16.7) %>%
    filter(role == "Combo G" | role == "Scoring PG"
        | role == "Wing G" | role == "Pure PG") %>%
  filter(!yr == "Sr") %>%
  select(player,team)

ggarrange(r1,r2, ncol = 2, nrow = 1)
```

## Assists

We filter down to the past 3 season's top 25 teams.

```{r top 10 asts, echo = F, message = F, fig.height = 3, fig.width = 8}
options(scipen = 999)
a1 = top_25s %>%
  na.omit() %>%
  filter(appearances == 3) %>%
  group_by(team) %>%
  summarize(n = sum(ast))%>%
  arrange(desc(n)) %>%
  ggplot()+
  geom_col(aes(x = team,y = n,fill = team))+
  ggtitle("Total Assist for Team with 3 appearances")+
  xlab("Team")+
  ylab("Assists")+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank())
  

some_new_bb = new_bb %>%
  filter(top_team == T) %>%
  left_join(new_ranks)

a2 = some_new_bb %>%
  group_by(team) %>%
  summarize(n_ast = mean(ast)) %>%
  left_join(some_new_bb) %>%
  na.omit() %>%
  ggplot() +
  geom_density(aes(x = n_ast, color = as.factor(appearances)))+
  ggtitle(label = "Mean Assists of All Teams Appearing in the Top 25",
        subtitle = "Independent of Season Rank For All Seasons")+
  labs(color = "Appearances in Top 25")

ggarrange(a1,a2, ncol = 2, nrow = 1)

#2-3 of the rank 1s for past 3 seasons appeared 3x
new_ranks %>%
  filter(rank == 1)
```

## The assist to turnover ratio 

The assist to turnover ratio is one of many basketball statistics used to evaluate the ball control and ball handling skills of a player


```{r ast.tov, message = F, echo = F, fig.height = 3, fig.width = 8}
# quantile(new_bb$ast.tov,c(.25,.5,.75), na.rm = T)

at1 = new_bb %>%
  filter(GP > 29) %>%
  ggplot(aes(x = GP,y = ast.tov)) +
  geom_point() + 
  geom_smooth()+ 
  xlab("GP")+
  ylab("Assist/Turnover ratio")+
  ggtitle(label = "Rebound Comparison",
          subtitle = "all players")

some_new_bb = new_bb %>%
  filter(top_team == T) %>%
  left_join(new_ranks) %>%
  na.omit()


at2 = some_new_bb %>%
  group_by(team) %>%
  summarize(at_tot = sum(round(ast.tov,2))) %>%
  left_join(some_new_bb) %>%
  ggplot() +
  geom_density(aes(x = at_tot, color = as.factor(appearances)))+
  ggtitle(label = "Sum of Assist/Turnover Ratio of All Teams Appearing \nin the Top 25",
        subtitle = "Independent of Season Rank For All Seasons")+
  labs(color = "Appearances in Top 25")+
  xlab("Assist/Turnover Ratio")

ggarrange(at1,at2, ncol = 2, nrow = 1)
```

## Conclusion
~~Get Buckets~~

Teams with high amounts of assists are getting high amounts of points.

Competitive players shoot free throws with around 75% accuracy.
The guy with the highest free throw accuracy kinda sucks.

Most people getting rebounds are post players

Made discoveries on how assists, while important do not necessarily prove a team's skill



## References
Source of some questions: 

https://www.breakthroughbasketball.com/stats/how-we-use-stats-Hagness.html

Data source:

https://www.kaggle.com/adityak2003/college-basketball-players-20092021/metadata



